hatchR: A toolset to predict hatch and emergence phenology in wild fishes

### 2024-12-06 README.Rmd
chnage `hatchR` to **hatchR**

`hatchR` is an R package for predicting fish developmental phenology using statistical models. It is intended to bridge the analytic gap of taking statistical models developed in hatchery settings (e.g., Beacham and Murray 1990) and applying them to real world temperature data sets using the effective value framework developed by Sparks et al. (2019).

@@

If you use `hatchR` in a formal publication or report, please cite it. Citing `hatchR` lets us devote more resources to it in the future. View the `hatchR` citation by running
```r
citation(package = "hatchR")
```

@@

You can install the development version of `hatchR` from [GitHub](https://github.com/) with:

@@

## Contributing to `hatchR`

We encourage users to report bugs and/or contribute to `hatchR`. For more detail on how to do this, please see our contributing guide (`CONTRIBUTING.md`).

@@

There are several ways to get help with `hatchR`:

@@

Below we provide a brief example showing how to use `hatchR`. For a thorough introduction to the software, see our introductory vignette [linked here](https://bmait101.github.io/hatchR/articles/Introduction.html). For a list of all functions available in `hatchR`, see our function reference [linked here](https://bmait101.github.io/hatchR/reference/index.html).

First we load `hatchR` and helper packages by running

@@

The example `woody_island` dataset is included in `hatchR` and contains temperature data from Woody Island, Lake Iliamna, Alaska. It includes daily water temperature data from 1990-1992. We can use `plot_check_temp()` to visually check the data:

@@

`hatchR` suggests the following **R** packages:

@@

`hatchR` project code is provided on an "as is" basis and the user assumes responsibility for its use. USDA Forest Service has relinquished control of the information and no longer has responsibility to protect the integrity , confidentiality, or availability of the information. Any reference to specific commercial products, processes, or services by service mark, trademark, manufacturer, or otherwise, does not constitute or imply their endorsement, recommendation or favoring by USDA Forest Service.

>>>>

**hatchR** is an R package for predicting fish developmental phenology using statistical models. It is intended to bridge the analytic gap of taking statistical models developed in hatchery settings (e.g., Beacham and Murray 1990) and applying them to real world temperature data sets using the effective value framework developed by Sparks et al. (2019).

@@

If you use **hatchR** in a formal publication or report, please cite it. Citing **hatchR** lets us devote more resources to it in the future. View the **hatchR** citation by running
```r
citation(package = "hatchR")
```

@@

You can install the development version of **hatchR** from [GitHub](https://github.com/) with:

@@

## Contributing to **hatchR**

We encourage users to report bugs and/or contribute to **hatchR**. For more detail on how to do this, please see our contributing guide (`CONTRIBUTING.md`).

@@

There are several ways to get help with **hatchR**:

@@

Below we provide a brief example showing how to use **hatchR**. For a thorough introduction to the software, see our introductory vignette [linked here](https://bmait101.github.io/hatchR/articles/Introduction.html). For a list of all functions available in **hatchR**, see our function reference [linked here](https://bmait101.github.io/hatchR/reference/index.html).

First we load **hatchR** and helper packages by running

@@

The example `woody_island` dataset is included in **hatchR** and contains temperature data from Woody Island, Lake Iliamna, Alaska. It includes daily water temperature data from 1990-1992. We can use `plot_check_temp()` to visually check the data:

@@

**hatchR** suggests the following **R** packages:

@@

**hatchR** project code is provided on an "as is" basis and the user assumes responsibility for its use. USDA Forest Service has relinquished control of the information and no longer has responsibility to protect the integrity , confidentiality, or availability of the information. Any reference to specific commercial products, processes, or services by service mark, trademark, manufacturer, or otherwise, does not constitute or imply their endorsement, recommendation or favoring by USDA Forest Service.

### 2024-12-06 inst/manuscript/paper.Rmd
chnage `hatchR` to **hatchR**

Understanding the timing of key life history events is necessary for managing and conserving populations. Historically, models to predict hatch and emergence timing for fishes were difficult to employ in wild settings because average incubation temperature was needed as the primary parameter in predictive models. However, recent improvements to these techniques reworked models such that they could be applied in wild environments as long as users had data for when adult fish spawned and a record of average daily temperature over the course of development. Despite these improvements, their application remains limited due to few parameterizations for varying species, being largely limited to salmonids. Here we present `hatchR`, a software ecosystem that allows users to predict hatch and emergence timing for wild fishes, as well as additional tools to aid in those analyses. `hatchR` allows users to leverage popular historic parameterizations for phenological models or to easily implement custom parameterizations using data not included in the package. `hatchR` is also distributed in two forms—an open source R package for maximum customization, as well as an HTML graphical-user-interface web application for individuals not familiar with scripting languages. To demonstrate potential uses, we present two case studies as likely applications for this software. `hatchR` promises to open many exciting avenues in research and management of fishes during their early life history.

@@

The resulting effective value approach has now been widely applied in Salmonids for which parameterizations from aquaculture were readily available—for example Pacific Salmon (*Oncorhynchus spp.*) models developed by @beacham1990 have been applied to various species and populations [@adelfio2024; @adelfio2019; @kaylor2021] while models developed for Bull Trout (*Salvelinus confluentus*) by @mcphail1979 were extended by @austin2019. Despite growing popularity, applications have been largely limited within Salmonids, presumably because parameterizations for such models already existed due to their wide use in aquaculture and their general popularity as sport and commercial fish.

To bridge the gap between the application of one-off effective value model applications within individual studies and the lack of parameterization for other species, we developed the software ecosystem, `hatchR`. Specifically, `hatchR` allows users to input standard raw or summarized temperature datasets that are commonly collected in wild settings, run basic checks on those data, use built-in parameterizations like those from @beacham1990 or @sparks2017, develop custom models from their own or published temperature and phenological data, and predict hatch and emergence timing using these models in the effective value framework.

To widen the user application of these methods, we distribute two user-interfaces for `hatchR`. The first is a R package distributed via CRAN that allows users the most customizable application for these methods. The R package is especially powerful as it allows users to automate their analyses over multiple variables such as phenology type, multiple spawn dates, or different habitats with varying thermal regimes. These variable approaches are outlined in the package documentation on `hatchR`'s website. Alternatively, we also distribute a Shiny application in the form of an HTML-based web tool to interact with many of `hatchR`'s functions in a graphical-user-interface. The Shiny form trades-off some of automative power for user simplicity, while still allowing users to leverage much of the functionality of `hatchR`'s R package. Below, we present the basic overview of the software and multiple case studies of how it may be applied.

```{r echo=FALSE}
### 
library(tidyverse); library(hatchR)
```

@@

`hatchR` is meant to primarily be a tool for predicting phenology. In this sense, we mostly limit functionality to these applications and provide minimal data checking and plotting help. This decision is in part driven by the diversity of data types that users may import and the difficulty in addressing all those data types with respect to various data checks. In other words, we expect users to know their data better than we do and to check it accordingly. We do provide two basic data check functions discussed in the Checking Data section. Similarly, we provide limited functionality for plotting results, but provide examples of how to build custom visualization from output, specifically in R. For the Shiny application, we provide a base output plot, but the ability to download your results for custom plotting in programs of the user's choice.

@@

Effective value models were created by @sparks2019 to implement developmental models in wild environments for Sockeye Salmon (*O. nerka*). The need for their development arose because historic models, specifically those in @beacham1990, only considered the average incubation temperature during development and, for wild fishes, average incubation temperature was impossible to estimate because it was unknown when fish hatched even if adult spawn timing was known. To address this, @sparks2019 used the reciprocal of the formulation of model 2 from @beacham1990 and assigned an effective value for every day of development using the daily average temperature.

@@

Where *i* is the daily value and a fish hatches or emerges when the cumulative sum reaches one: $$\sum_{i =1}^nEffectiveValue_i = 1$$

The effective value model framework is the basis for the phenological models in `hatchR`, both in the included `model_table` in the package (though the model table includes more complex models developed by @beacham1990) as well as for custom models users can fit with `fit_model()`. Specifically, `model_table` has been extended to include more parameterizations from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979).

@@

`hatchR` is built assuming data will be analyzed as daily average temperatures. Despite that assumption, raw data (*e.g.*, as outputted by HOBO loggers) can be used and `hatchR` includes functionality to summarize those data into a format that is usable, as well as provides functions for basic visual and programmatic data checks to make sure outliers or missing data are at least brought to users' attention.

We demonstrate the utility of the summary and check functions `summarize_temp()`, `plot_check_temp()`, and `check_continuous()` using a simulated year-long dataset with temperature readings every thirty minutes.

```{r}
# create date object for a year with 30 min reading intervals
dates <- seq(from = ymd_hms("2000-07-01 00:00:00"),
             to = ymd_hms("2001-06-30 23:59:59"), length.out = 17568)

# create empty dataframe
year_sim <- data.frame(matrix(NA, nrow = length(dates), ncol = 1))

# date column
colnames(year_sim)[1] <- "date"

# add dates vector to date column
year_sim[1] <- dates

#random seed
set.seed(123)

# take temps from a random normal dist with mean 10 sd 3 
# for every date time combo in dates and append to column (temp) in year_sim
year_sim$temp <- rnorm(n = length(dates), mean = 10, sd = 3) %>%
  abs()

dim(year_sim)

First, we recommend checking import data for any outliers or strange inputs using `plot_check_temp()`

```{r}
plot_check_temp(data = year_sim,
                dates = date,
                temperature = temp,
                temp_min = 0, # temp_min and max lines are user customizable
                temp_max = 25)
```

There are no obvious outliers but since each day has 48 records, we need to summarize it to daily mean temperature with `summarize_temp()` and then check for missing days with `check_continuous()`. We also recommend using `plot_check_temp()` again on the summarized data.

```{r}
# summarize
year_sim_summ <- summarize_temp(data = year_sim,
                                dates = date,
                                temperature = temp)

dim(year_sim_summ)

# check continuous
check_continuous(data = year_sim_summ,
                 dates = date)

# we can demonstrate an error by removing the Oct 8 (100th day)
check_continuous(data = year_sim_summ[-100,],
                 dates = date)

# it is useful to plot again to check your summarized data
plot_check_temp(data = year_sim_summ,
                dates = date,
                temperature = daily_temp,
                temp_min = 0, # temp_min and max lines are 
                              # user customizable
                temp_max = 15)
```

@@

fit_model for three non-salmonid specieslib

Below, we demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids in the `model_table` included in the package. We include parameterizations from three warm-water species to demonstrate the `fit_model()` utility for species beyond the scope of the original effective value approach. We include parameterizations for commonly cultured sportfishes including Smallmouth Bass (*Micropterus dolomieu*), Channel Catfish (*Ictalurus punctatus*) from @small2001, and Lake Sturgeon (*Acipenser fulvescens*) from @smith2005.

We demonstrate the utility of this approach by creating a random thermal regime with an ascending thermograph with a mean temperature of 16 °C, parameterizing models for each species, and demonstrating days to hatch and developmental period for each species with the random thermal regime.

```{r}
###  make temp regime
set.seed(123)
# create random temps and corresponding dates
temps_sim <- sort(rnorm(n =30, mean = 16, sd = 1), decreasing = FALSE)
dates_sim <-  seq(from = ymd("2000-07-01"),
             to = ymd("2000-07-31"), length.out = 30)

data_sim <- matrix(NA, 30, 2) |> data.frame()
data_sim[,1] <- temps_sim
data_sim[,2] <- dates_sim

# change names so they aren't the same as the vector objects
colnames(data_sim) <- c("temp_sim", "date_sim")

### smallmouth mod
smallmouth <- matrix(NA, 10, 2) |> data.frame()
colnames(smallmouth) <- c("hours", "temp_F")
smallmouth$hours <- c(52, 54, 70, 78, 90, 98, 150, 167, 238, 234)
smallmouth$temp_F <- c(77, 75, 71, 70, 67, 65, 60, 59, 55, 55)

# change F to C and hours to days
smallmouth <- smallmouth |> 
  mutate(days = ceiling(hours/24),
         temp_C = (temp_F -32) * (5/9))

smb_mod <- fit_model(temp = smallmouth$temp_C,
                     days = smallmouth$days,
                     species = "smb",
                     dev.type = "hatch")

### catfish mod
catfish <- matrix(NA, 3, 2) |> data.frame()
colnames(catfish) <- c("days", "temp_C")
catfish$days <- c(16,21,26)
catfish$temp_C <- c(22,10,7)

cat_mod <- fit_model(temp = catfish$temp_C,
                     days = catfish$days,
                     species = "catfish",
                     dev.type = "hatch")

### lake sturgeon mod
sturgeon <-  matrix(NA, 7, 2) |> data.frame()
colnames(sturgeon) <- c("days", "CTU")
sturgeon$days <- c(7,5,6,6,5,11,7)
sturgeon$CTU <- c(58.1, 62.2, 61.1, 57.5, 58.1, 71.4, 54.7)

sturgeon <- sturgeon |> 
  mutate(temp_C = CTU/days) # change CTUs to average temp and add column

sturgeon_mod <- fit_model(days = sturgeon$days,
                          temp = sturgeon$temp_C,
                          species = "sturgeon",
                          dev.type = "hatch")

@@

Note the model the *R^2^* fit from the models below. You can see the generally all preform well and are in line with values from model 2 of @beacham1990.

```{r}
#model fits
smb_mod$r_squared; cat_mod$r_squared; sturgeon_mod$r_squared

We additionally provide the model estimates for hatch timing for each of the three species used to generate custom models in Figure XXX.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
### predict_phenology

#smallmouth bass
smb_res <- predict_phenology(data = data_sim,
                  dates = date_sim,
                  temperature = temp_sim,
                  spawn.date = "2000-07-01",
                  model = smb_mod$func)

# catfish
catfish_res <- predict_phenology(data = data_sim,
                  dates = date_sim,
                  temperature = temp_sim,
                  spawn.date = "2000-07-01",
                  model = cat_mod$func)

# sturgeon
# note that 16 C is pretty far out of range of temps for model fit, not best practice
sturgeon_res <- predict_phenology(data = data_sim,
                  dates = date_sim,
                  temperature = temp_sim,
                  spawn.date = "2000-07-01",
                  model = sturgeon_mod$func)

# summary for all species
all_res <- data.frame(matrix(NA, 3, 2))
colnames(all_res) <- c("start", "stop")

all_res$start <- c(catfish_res$dev.period$start, 
                   smb_res$dev.period$start, 
                   sturgeon_res$dev.period$start)

all_res$stop <- c(catfish_res$dev.period$stop,
                  smb_res$dev.period$stop, 
                  sturgeon_res$dev.period$stop)

all_res <- all_res |> 
  mutate(days = ceiling(stop-start),
         index = c(17,16.5,16))

all_res$Species <- c("Channel Catfish", "Smallmouth Bass", "Lake Sturgeon")

ggplot() +
  geom_point(data = data_sim, aes(x = date_sim, y = temp_sim )) + 
  geom_line(data = data_sim, aes(x = date_sim, y = temp_sim )) +
  geom_rect(data = all_res, aes(xmin = start, xmax = stop, ymax =index-.35, ymin = index-.5, fill = Species)) +
  geom_label(data = all_res, aes(x = start + (stop - start) / 1.25, y = (index -0.425), label = days)) +
  labs(x = "Date", y = "Temperature (°C)") +
  scale_fill_manual(values = c("deepskyblue4", "grey23", "darkolivegreen4")) +
  theme_classic() +
  theme(legend.position = c(0.75, 0.25))
```

@@

Use woody example from website

@@

In this scenario, we will consider the grazing example and Bull Trout, a threatened fish in the United States under the Endangered Species Act [@nolfi2024], and the East Fork Salmon River, a key Bull Trout population in the upper Salmon River watershed. The fisheries manager there wants to know if fish will likely be out of the gravel and free-swimming by June 1st. In this system, it is expected that Bull Trout will be done spawning by the end of September, so we'll consider the last possible spawn date as September 30th.

We demonstrate this first case study using the graphical user interface portion of the `hatchR` ecosystem found at <https://elifelts.shinyapps.io/hatchR/>. Users will first upload their data with the `Import Data` window, which requires them to select their file on their personal computer, provide the program with the columns corresponding for dates and temperatures, and then provide the format in which dates are coded (*e.g.*, year-month-day or day-month-year). Once data is uploaded the program automatically plots the user's data using `plot_check_temp()` in the background and provides them the outputted graphical check. After uploading and checking data, the user switches to the `Model Phenology` window. In this circumstance, we use the preloaded parameterization for bull trout from @austin2019 with the `Existing` button for model selection, which the user selects with the various drop down options in the menu. After the model is selected, the user can choose multiple spawn dates from the interactive calendar provided. We show results for spawning for September 30th as indicated in the example above. Once dates are chosen, a table entry for each spawn date is outputted in the `Phenology Summaries` tab and corresponding plot with data from each spawn date in the `Timeline Plot` tab. Output from predicting phenology and the resulting figure are downloadable from their respective tabs. The process is demonstrated in full in Figure XXX, but the interface is described more completely in the Articles on `hatchR`'s website <https://bmait101.github.io/hatchR/>.

```{r, eval=FALSE, echo=FALSE}

@@

We thank Laura Koller for her help designing the `hatchR` logo. Dan Isaak provided useful discussion about model development and temperature datasets.

>>>>

Understanding the timing of key life history events is necessary for managing and conserving populations. Historically, models to predict hatch and emergence timing for fishes were difficult to employ in wild settings because average incubation temperature was needed as the primary parameter in predictive models. However, recent improvements to these techniques reworked models such that they could be applied in wild environments as long as users had data for when adult fish spawned and a record of average daily temperature over the course of development. Despite these improvements, their application remains limited due to few parameterizations for different species, being largely limited to salmonids. Here we present **hatchR**, a software ecosystem that allows users to predict hatch and emergence timing for wild fishes, as well as additional tools to aid in those analyses. **hatchR** allows users to leverage popular historic parameterizations for phenological models or to easily implement custom parameterizations using data not included in the package. **hatchR** is also distributed in two forms—an open source R package for maximum customization, as well as an HTML graphical-user-interface web application for individuals not familiar with scripting languages. To demonstrate potential uses, we present two case studies as likely applications for this software. **hatchR** promises to open many exciting avenues in research and management of fishes during their early life history.

@@

The resulting effective value approach has now been widely applied in salmonids for which parameterizations from aquaculture were readily available—for example Pacific Salmon (*Oncorhynchus spp.*) models developed by @beacham1990 were have been applied to various species and populaions [@adelfio2024; @adelfio2019; @kaylor2021] while models developed for Bull Trout (*Salvelinus confluentus*) by @mcphail1979 were extended by @austin2019. Despite growing popularity, applications have been largely limited within Salmonids, presumably because parameterizations for such models already existed due to their wide use in aquaculture and their general popularity as sport and commercial fish.

To bridge the gap between the application of one-off effective value model applications within individual studies and the lack of parameterization for other species, we developed the software ecosystem, **hatchR**. Specifically, **hatchR** allows users to input standard raw or summarized temperature datasets that are commonly collected in wild settings, run basic checks on those data, use built-in parameterizations like those from @beacham1990 or @sparks2017, develop custom models from their own temperature and phenological data, and predict hatch and emergence timing using these models in the effective value framework.

To widen the user application of these methods, we distribute two user-interfaces for **hatchR**. The first is a R package distributed via CRAN that allows users the most customizable application of these methods. The R package is especially powerful as it allows users to automate their analyses over multiple variables such as phenology type, multiple spawn dates, or different habitats with varying thermal regimes. These variable approaches are outlined in the package documentation on **hatchR**'s website. Alternatively, we also distribute a Shiny application in the form of an HTML-based web tool to interact with many of **hatchR**'s functions in a graphical-user-interface. The Shiny form trades-off some of automative power for user simplicity, while still allowing users to leverage much of the functionality of **hatchR**'s R package. Below, we present the basic overview of the software and multiple case studies of how it may be applied.

@@

**hatchR** is meant to primarily be a tool for predicting phenology. In this sense, we mostly limit functionality to these applications and provide minimal data checking and plotting help. This decision is in part driven by the diversity of data types that users may import and the difficulty in addressing all those data types with respect to various data checks. In other words, we expect users to know their data better than we do and to check it accordingly. We do provide two basic data check functions discussed in the Checking Data section. Similarly, we provide limited functionality for plotting results, but provide examples of how to build custom visualization from output, specifically in R. For the Shiny application, we provide a base output plot, but the ability to download your results for custom plotting in programs of the user's choice.

@@

Effective value models were created by @sparks2019 to implement developmental models in wild environments for Sockeye Salmon (*O. nerka*). The need for change arose because historic models, specifically those in @beacham1990, only considered the average incubation temperature during development and, for wild fishes, average incubation temperature was impossible to estimate because it was unknown when fish hatched even if adult spawn timing was known. To address this, @sparks2019 used the reciprocal of the formulation of model 2 from @beacham1990 and assigned an effective value for every day of development using the daily average temperature.

@@

Where *i* is the daily value and a fish hatches or emerges when the cumulative sum or $$\sum_{i =1}^nEffectiveValue_i = 1$$

The effective value model framework is the basis for the phenological models in **hatchR**, both in the included `model_table` in the package, as well as for custom models users can fit with `fit_model()`. Specifically, `model_table` has been extended to include more parameterizations from @beacham1990, @sparks2017, and @austin2019 (who extended @mcphail1979).

@@

**hatchR** is built assuming data will be analyzed as daily average temperatures. Despite that assumption, raw data (*e.g.*, as outputted by HOBO loggers) can be used and **hatchR** includes functionality to summarize those data into a format that is usable, as well as provides functions for basic visual and programmatic data checks to make sure outliers or missing data are at least brought to users' attention.

summarize_temp, check_continuous and plot_check_temp

@@

fit_model for three non-salmonid species

```{r, eval=FALSE}
smallmouth <- matrix(NA, 10, 2) |> tibble()
colnames(smallmouth) <- c("hours", "temp_F")
smallmouth$hours <- c(52, 54, 70, 78, 90, 98, 150, 167, 238, 234)
smallmouth$temp_F <- c(77, 75, 71, 70, 67, 65, 60, 59, 55, 55)

smallmouth <- smallmouth |> 
  mutate(days = ceiling(hours/24),
         temp_C = (temp_F -32) * (5/9))

plot(smallmouth$days~smallmouth$temp_C)

smb_mod <- fit_model(smallmouth,
                     temp = smallmouth$temp_C,
                     days = smallmouth$hours)

@@

@@

Use woody \_example from website

@@

In this scenario, we will consider the grazing example and Bull Trout, a threatened fish in the united states under the Endangered Species Act [@nolfi2024], and the East Fork Salmon River, a key Bull Trout population in the upper Salmon River watershed. The fisheries manager there wants to know if fish will likely be out of the gravel and free-swimming by June 1st. In this system, it is expected that Bull Trout will be done spawning by the end of September, so we'll consider the last possible spawn date as September 30th.

```{r}

@@

We would like to sincerely thank the editor and reviewers for all of their helpful feedback which greatly improved both the software and the manuscript.
